Effectful Semantics for Substructural SSA

### 2023-11-13 pldi24/paper.tex
Tap

We define a \textbf{program order pomset} to be a pomset with \(\mc{A} =
\mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
\textit{reads} of the form \(x = v\) and \(\mc{A}_w\) consists of
\textit{writes} of the form \(x := v\) for locations \(x\) and values \(v \in
\ms{Word}\). We define the \textbf{program order monad} \(\ms{PO}\;A =
\ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is the stream action on finite
program order pomsets given by sequential composition for \(\mc{A}\). This
yields an \isotopessa model with support for concurrent read and write
operations \(\ms{read}_x: \mc{I}^\varnothing_0(\mb{1}, \ms{Word})\),
\(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with
semantics
\begin{equation}
  \dnt{\ms{read}_x} = \lambda (). \{(v, \{x = v\}) | v \in \ms{Word}\}
  \qquad
  \dnt{\ms{write}_x} = \lambda v. \{((), \{x := v\})\}
\end{equation}
The semantics of \(\ms{read}\) in particular give a hint as to how this model
works: rather than tracking the state of the heap, we simply \textit{emit} a
pomset of the events that would have to happen for a given execution (in the
case of a read, a read event \(x = v\) whenever the read returns \(v\), and in
the case of a write, the single write event \(x := v\) for a write of \(v\)),
and later post-filter on the pomset of an entire program to get a sequentially
consistent execution. Hence, our semantics remains compositional, allowing us to
reason about each program fragment individually, while still allowing us to add
arbitrary external state (e.g. another program concurrently modifying the heap)
before finally considering the execution of an entire program.

On that note, we support the parallel execution of multiple (potentially
nonterminating) morphisms as follows:
\begin{equation}
  \begin{aligned}
  f || g = \lambda (a, b). 
  & \{\iota_0 ((a', b'), \alpha || \beta) 
    \mid \iota_0 (a', \alpha) \in f\;a 
    \land \iota_0 (b', \beta) \in f\;b\} 
  \\ & \cup \{\iota_1 (\alpha || \beta) \mid (\iota_0 (a', \alpha) \in f\;a \lor \iota_1\;\alpha \in f\;a) \land \iota_1 \beta \in f\;b\} 
  \\ & \cup \{\iota_1 (\alpha || \beta) \mid \iota_1\;\alpha \in f\;a \land \iota_0 (b', \beta) \in f\;b\}: \ms{Set}_{\ms{PO}}(A \otimes B, A' \otimes B')
  \end{aligned}
\end{equation}
It's tempting to try to use this as a tensor product of morphisms and so
graduate to a bona fide monoidal category rather than our current premonoidal
setting, since spatial concurrency is one of the headline use-cases of the
former, but this definition does \textit{not} support sliding:
\begin{equation}
  (\ms{write}_x || \ms{id}) ; (\ms{id} || \ms{write}_y)
  \neq (\ms{id} || \ms{write}_y) ; (\ms{write}_x || \ms{id})
\end{equation}
since the left-hand side will emit the pomset \([x := a, y := b]\) while the right-hand side will emit the pomset \([y := b, x := a]\). If we did want this behaviour, we'd need to have a significantly more sophisticated model of composition which effectively takes nontermination into account, which we will leave to future work.

To graduate from SC concurrency to a genuine, if maximally simple, weak memory
model, we need to introduce a \textit{buffer} of type \(\ms{Buf} = \mc{A}^*\) of
lists of actions, which we will interpret as linear pomsets ordered by index. In particular, we define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;\ms{PO}\), which gives us another \isotopessa model in the usual manner.

@@

@@

\TODO{too powerful, submonad rune of buffer continuity}

>>>>

We define a \textbf{program order pomset} to be a pomset with \(\mc{A} =
\mc{A}_w \cup \mc{A}_r \cup \{\delta\}\), where \(\mc{A}_r\) consists of
\textit{reads} of the form \(x = v\) and \(\mc{A}_w\) consists of
\textit{writes} of the form \(x := v\) for locations \(x\) and values \(v \in
\ms{Word}\). We define the \textbf{program order monad} \(\ms{PO}\;A =
\ms{Traces}\;\Sigma\;A\), where \(\Sigma\) is the stream action on finite
program order pomsets given by sequential composition for \(\mc{A}\). This
yields an \isotopessa model with support for concurrent read and write
operations \(\ms{read}_x: \mc{I}^\varnothing_0(\mb{1}, \ms{Word})\),
\(\ms{write}_x: \mc{I}^\varnothing_0(\ms{Word}, \mb{1})\) with
semantics
\begin{equation}
  \dnt{\ms{read}_x} = R_x^{\ms{PO}} = \lambda (). \{(v, \{x = v\}) | v \in \ms{Word}\}
  \qquad
  \dnt{\ms{write}_x} = W_x^{\ms{PO}} =  \lambda v. \{((), \{x := v\})\}
\end{equation}
The semantics of \(\ms{read}\) in particular give a hint as to how this model
works: rather than tracking the state of the heap, we simply \textit{emit} a
pomset of the events that would have to happen for a given execution (in the
case of a read, a read event \(x = v\) whenever the read returns \(v\), and in
the case of a write, the single write event \(x := v\) for a write of \(v\)),
and later post-filter on the pomset of an entire program to get a sequentially
consistent execution. Hence, our semantics remains compositional, allowing us to
reason about each program fragment individually, while still allowing us to add
arbitrary external state (e.g. another program concurrently modifying the heap)
before finally considering the execution of an entire program.

On that note, we support the parallel execution of multiple (potentially
nonterminating) morphisms as follows:
\begin{equation}
  \begin{aligned}
  f || g = \lambda (a, b). 
  & \{\iota_0 ((a', b'), \alpha || \beta) 
    \mid \iota_0 (a', \alpha) \in f\;a 
    \land \iota_0 (b', \beta) \in f\;b\} 
  \\ & \cup \{\iota_1 (\alpha || \beta) \mid (\iota_0 (a', \alpha) \in f\;a \lor \iota_1\;\alpha \in f\;a) \land \iota_1 \beta \in f\;b\} 
  \\ & \cup \{\iota_1 (\alpha || \beta) \mid \iota_1\;\alpha \in f\;a \land \iota_0 (b', \beta) \in f\;b\}: \ms{Set}_{\ms{PO}}(A \otimes B, A' \otimes B')
  \end{aligned}
\end{equation}
It's tempting to try to use this as a tensor product of morphisms and so
graduate to a bona fide monoidal category rather than our current premonoidal
setting, since spatial concurrency is one of the headline use-cases of the
former, but unfortunately, sliding still fails:
\(
  (W_x^{\ms{PO}} || \ms{id}) ; (\ms{id} || W_y^{\ms{PO}})
  \neq (\ms{id} || W_y^{\ms{PO}}) ; (W_x^{\ms{PO}} || \ms{id})
\),
since the left-hand side will emit the pomset \([x := a, y := b]\) while the
right-hand side will emit the pomset \([y := b, x := a]\). If we did want this
behaviour, we'd need to have a significantly more sophisticated model of
composition which effectively takes nontermination into account, which we will
leave to future work.

To graduate from SC concurrency to a genuine, if maximally simple, weak memory
model, we need to introduce a \textit{buffer} of type \(\ms{Buf} = \mc{A}_w^*\)
of lists of write actions, which we will interpret as linear pomsets ordered by
index. In particular, we define the monad \(\ms{TSO} =
\ms{StateT}\;\ms{Buf}\;\ms{PO}\). Given a buffer \(\ms{Buf}\), we can define the
result of reads \([\cdot]_x: \ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the
buffer as follows:
\begin{equation}
  (L;\{x := v\})[x] = v
  \qquad
  (L;\{y := v\})[x] = L[x]
  \qquad
  [][x] = \bot
\end{equation}
Note that writes at the \textit{end} of the buffer are prioritized, since later
writes overwrite earlier ones! We will implement TSO ordering by buffering all
our writes, in which case they are only visible to the local thread. At any
point, we may choose to \textit{flush} some of the writes in buffer to the
global pomset of events. This is implemented via the morphism (called
\(\ms{split}\) in \cite{sparky})
\begin{equation}
  \ms{pflush} = \lambda ()\;L. \{((), L', L'') | L = L';L''\}
  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
\end{equation}
On the other hand, reads events will first attempt to read from the buffer, and,
if there is no corresponding write in the buffer, will read an arbitrary value,
in both cases pushing an event to the global pomset; that is, the semantics of
reads and writes are given by
\begin{equation}
  \begin{aligned}
  \dnt{\ms{read}_x} = R_x^{\ms{TSO}} 
    &= \ms{pflush};(\lambda ()\;L. \{(v, L, \{x = v\}) \mid L[x] = v \lor L[x] = \bot\});\ms{pflush} \\
  \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
    &= \ms{pflush};(\lambda v\;L. \{(v, L)\});\ms{pflush}
  \end{aligned}
\end{equation}

@@

\TODO{in the presence of parallel composition}

@@

\TODO{big power, such transactional memory, much wow}
